Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-17 writeup/pnas-word-order-si.tex
code

Here we replicate Study 1 with both parsers and language models operating entirely on word forms, without POS tags.
Results are shown in Figures~\ref{fig:pareto-plane-noPOS}.
As shown in Figure~\ref{fig:byLambda-noPOS}, most real grammars have higher efficiency than most baselines across permissible values of $\lambda$.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}\label{fig:pareto-plane-noPOS}
\end{figure}

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized.tex}
\end{tabular}
}
	\caption{Study 1, replication with lexicalized models: Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}\label{tab:pareto-plane-noPOS}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../results/plane/unlexicalized/analyze_pareto_optimality/figures/quantileByLambda-mle.pdf}
	\caption{Study 1, replication with lexicalized models: Optimality by $\lambda$ (lexicalized models).}\label{fig:byLambda-noPOS}
\end{figure}

@@

In Study 1, we considered deterministic ordering grammars, even though real languages have some degree of word order freedom.
Here, we show that deterministic baseline grammars provide conservative baselines: They have higher efficiency than baseline grammars with word order freedom comparable to real languages.

In order to obtain baselines whose freedom is comparable to that of real languages, we constructed baselines that have the same Branching Direction Entropy \cite{futrell2015quantifying}:
For every one of the 37 syntactic relations, we chose $a_\tau$ the direction entropy to be the same as observed in the actual orderings found in the UD corpus.
We kept the relative ordering of siblings on one side deterministic, i.e., we considered the limit where the values $b_\tau$ for different $\tau$ are far apart.

Results are shown in Figure~\ref{fig:nondeterministic-english}.
For every one of the baseline grammars, we show both its deterministic and its nondeterministic version.
Nondeterministic grammars are strongly less efficient than deterministic grammars, showing that deterministic grammars constitute stronger baselines.

\begin{figure}
	\centering
	\includegraphics[width=0.4\textwidth]{../results/plane/nondeterministic/pareto-plane-perLanguage-nondeterministic-mle.pdf}
	\caption{Parseability and predictability, including nondeterministic versions of the 50 baseline grammars.}\label{fig:nondeterministic-english}
\end{figure}

>>>>

In Study 1, we calculate parseability on the part-of-speech level, and also add part-of-speech tags when calculating predictability.
These choices are intended to prevent early overfitting during the grammar optimization process (Section~\ref{sec:neural-architectures}).
However, such unlexicalized parsers are less accurate than parsers taking acual word-forms into account, and adding part-of-speech tags might provide additional disambiguation that is absent in the original word-level input.
Here, we show that these limitations do not affect conclusions from Study 1, by replicating Study 1 with both parsers and language models operating entirely on word forms, without POS tags.
Results are shown in Figure~\ref{fig:pareto-plane-noPOS} and Table~\ref{tab:pareto-plane-noPOS}. In agreement with the previous results (Figure~\ref{fig:pareto-per-lang}), real grammars are mostly to the top right of their corresponding baselines.
We further confirm this in Figure~\ref{fig:lambda-quantile-noPOS}, which shows that most real grammars have higher efficiency than most baselines across permissible values of $\lambda$.
In fact, comparing Figure~\ref{fig:lambda-quantile-noPOS} to Figure~\ref{fig:lambda-quantile} suggests that optimality of real grammars is \emph{more} pronounced when modeling predictability and parseability fully on the level of word forms.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized-mle.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}\label{fig:pareto-plane-noPOS}
\end{figure}

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized-ground.tex}
\end{tabular}
}
	\caption{Study 1, replication with lexicalized models: Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}\label{tab:pareto-plane-noPOS}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../results/plane/unlexicalized/analyze_pareto_optimality/figures/quantileByLambda-mle.pdf}
	\caption{Study 1, replication with lexicalized models: Optimality of real grammars for efficiency, compared to baselines, across values of $\lambda$: The $x$-axis shows $\lambda \in [0,1)$, the $y$-axis shows the fraction of baselines that have lower efficiency than the real grammar at this value of $\lambda$, with 95\% confidence bands obtained from a two-sided binomial test. Compare Figure~\ref{fig:lambda-quantile}.}\label{fig:lambda-quantile-noPOS}
\end{figure}

@@

In Study 1, we considered deterministic ordering grammars, even though real languages have some degree of word order freedom.
Here, we show that deterministic baseline grammars provide conservative baselines: They have higher efficiency than baseline grammars with word order freedom comparable to the orders found in the original corpora, and thus provide conservative baselines for comparison with other grammars.

In order to obtain baselines whose freedom is comparable to that of real languages, we constructed baselines that have the same Branching Direction Entropy \cite{futrell2015quantifying} as observed in the original corpora.
The Branching Direction Entropy measures the extent of freedom in choice between head-final and head-initial orderings, and it is a corpus-based quantitative measure of word order freedom \cite{futrell2015quantifying}.
For a given syntactic relation, the \emph{branching direction entropy} measures the entropy of the Bernoulli variable that is $1$ whenever the head is ordered before the dependent, and $0$ if the dependent is ordered before the head.
The branching direction entropy is $0$ if only one of the two orders can occur, and it is $\log 2$ if both orders are equally frequent.

We constructed baseline grammars that match the branching direction entropies found in the original orders found in the corpora.
To this end, we converted the baseline grammars into differentiable ordering grammars (Section~\ref{sec:diff-gramm}).
Such grammars have parameters $a_\tau, b_\tau$ for each relation $\tau$.
For every one of the 37 syntactic relations, we chose $a_\tau$ so as to match the the direction entropy to that observed in the actual orderings found in the UD corpus.
For $b_\tau$, we considered the limit where the values $b_\tau$ for different relations $\tau$ are very far apart, making the relative ordering of siblings on the same side of the head fully deterministic.
That is, these ordering grammars match word order freedom as quantified by Branching Direction Entropy, and show no additional degrees of order freedom.

We compared these nondeterministic baseline grammars to their deterministic versions, for one language with relatively free order (Czech), and for two languages with relatively fixed order (English and Japanese).
Results are shown in Figure~\ref{fig:nondeterministic}.
For every one of the baseline grammars, we show both its deterministic and its nondeterministic version.
Nondeterministic grammars are less efficient than deterministic grammars, in particular in language with greater degrees of word order freedom (Czech).
This shows that deterministic grammars constitute stronger baselines.

\begin{figure}
	\centering
	\includegraphics[width=0.95\textwidth]{../results/plane/nondeterministic/pareto-plane-perLanguage-nondeterministic-mle.pdf}
	\caption{Parseability and predictability for three languages, including both deterministic (green) and nondeterministic  (blue) versions of the 50 baseline grammars.}\label{fig:nondeterministic}
\end{figure}
